How has the Brazilian Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2023-01-31 Submission/Re-submission/draft/draft_resubmission.rmd
Updated text and plots to be consistent with new changes

## 3.1 Codebook (Amazonian historiography)

To capture the cultural dimension and historical contexts behind problem constructions, we rely on the Amazonian historiography literature. We refer to Amazonian historiography as the body of research conducted by social and environmental scientists that tell the story of diverse governmental programs that studies the Amazon region, forest, and its people. Within this literature, we identify three historical-political constructions: national sovereignty, economic integration, and environmental conservation. We argue a fourth separate one also exists: social development. The social development construction focuses on the constitutional rights and dignity of the inhabitants of the Amazonian region.

@@

We code Amazonian statements in three steps. First, we develop a codebook for coding Amazonian problem-constructions identified with the Amazonian historiography (see appendix). In their conceptualization, each problem-construction is mutually exclusive; that is, they cover different forms of constructing the Amazon as a problem. However, an Amazonian statement can be coded as referring to multiple problem constructions (see Figure 1 below). Amazonian statements, thus, can be pure-type statements (i.e. constructs the Amazon as a single problem) or mixed-types (i.e. constructs the Amazon as multiple issues).

@@

To analyze our data, we model the share of a specific pure problem-construction, as the dependent variable, relative to the settings in which a particular speech happens, as our independent variable. This means we predict the probability of a president constructing the Amazon as an issue of (1) environmental conservation, (2) economic integration, (3) social development, and (4) national sovereignty, separately, and in relation to the remaining problem-constructions (see table 1 below). We focus on single problem constructions to provide unambiguous interpretations and more robust findings. We divide settings in terms of non-Amazonian Brazilian states, Amazonian states within Brazil, Brasilia, Amazonian countries, and international. This level of aggregation refers to geo-political settings that make sense in light of the Amazon, as our policy object. This aggregation combines different expectations presidents can have about transnational audiences in each of these settings. Brasilia, specifically, is treated as a separate setting due to its unique location and since most presidential speeches are delivered there. The model controls for yearly deforestation rates, inflation, election years, and peak years for Amazon in presidential speeches.

@@

Second, our dataset covers only what is considered an official speech. Presidents, though, give interviews, appear in debates, talk at campaign rallies, and, more recently, post on social media. Problem-construction within presidential discourse, thus, happens in different public, private, and online settings for which we do not account or differentiate in this paper. Furthermore, we aggregate levels to socio-political-cultural settings that make sense in light of the policy object in question and the data at hand. Therefore, we are aware that a category as "international", for instance, encompasses speeches delivered by presidents in a wide variety of countries, institutions, and venues to diverse audiences, which are not differentiated.

Third, as presidents participate and shape transnational politics, we assume that presidential discourse, policies, and outcomes are co-constitutive. Discourses capture general problem constructions that can be used to introduce, discuss, and/or justify policy and outcomes. Throughout the analysis we carefully discuss how presidential discourses relate to, for instance, deforestation rates and political outcomes by pointing at some of the patterns and correlations but without making any causal claims.

@@

We subsequently observe a steady increase reaching 24% in 2019, the first year of Bolsonaro's presidency. At the time, international and national media brought attention to record burning of the Amazon, which led up to a red sky afternoon in São Paulo in August 2019. Alongside this, Bolsonaro's dismantling of environmental governance and the threats to leave the Paris Agreement had strong media attention, and, as a response, the topic became prominent in his speeches. Bolsonaro went on to retrieve Brazil's hosting status for COP25. Finally, and unfortunately, in 2022, environmental activist, Bruno Mendes, and British journalist, Dom Phillips, were murdered and buried in the Amazon (McCoy 2022).

@@

Although we observe variation in the frequency of the Amazon as a general topic, our conceptual framework foresees variation in specific problem-construction depending on a myriad of factors. In this subsection, we investigate how pure-types and mixed-types problem constructions evolved over time.The same president simultaneously constructs the Amazon as various problems, which unravels a less cohesive story than that told by the policy-cycle literature. Nevertheless, the overall balance between problem-constructions reveals longitudinal trends that speak to the importance of the environment, economy, social policy, and sovereignty. Figure 3, below, illustrates the share of each pure-type problem-constructions in time as these are the most common.

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso's presidency (1995-2002), when economic integration was increasing while environmental conservation decreased. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. While this correlates with strong environmental policies being adopted in 2004, economic integration still ranked first. It was only in 2010, with President Dilma, that environmental conservation and social development problem-constructions surpassed economic integration. Interestingly, there is a steady increase of sovereignty problem-constructions starting in 2008 already. This reversal was not long-standing, though. With the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration increases to become the most favored construction.These trends reveal a multi-faceted relationship between problem-constructions, which we summarize in three findings.

@@

First, the same governments deploy contradictory constructions, but their relative frequencies reveal trends over time. The 1988 Brazilian Constitution foresaw the strengthening of environmentalism through the participation of indigenous peoples and civil society in policy-making [@keck1998], which was slowly confirmed, at least discursively. The fall of economic integration problem-construction and the rise of both social development and environmental conservation constructions in the 2000s suggest an unprecedented balance between granting local livelihoods their rights and economic development: never, since 1985, Amazonian constructions that focused on the environment and the constitutional rights of its people were more present than the economic integration in presidential discourses. While social development and environmental conservation did go down, we are now at a point in time where they appear almost as frequently as economic integration.

Second, the steady increase of sovereignty problem-constructions starting in 2008 offers context for the Presidency of Bolsonaro. The revision of the Forest Code, adopted in 2012, inaugurated a new phase of strong representation of agri-businesses in government and congress and, consequently, strong civil society criticism of government. In a period when sovereignty-related issues broadened from nation-states to a wider set of actors, stronger criticism of civil society in politics was often met by presidents with accusations of internationalization attempts. Thus, the political forces in Brazilian politics that drove policy and discursive changes in problem-construction were long in the making. Bolsonaro's problem-constructions, characterized by a unique combination of economic integration and sovereignty, is the strongest form of this shift.

<!-- Finally, our operationalization foresees the possibility of presidents might mixing multiple problem-constructions within a same statement about the Amazon. Mixed-type problem-constructions in discourse offer more intricate understandings of the Amazon as a problem. Constructing the Amazon as multiple issues averages at 18% of all constructions over time. Some mixed-types rarely appear so we focus our discussion on those mixed-types with higher incidence. Figure 4, below, displays the averages of mixed-types constructions by presidents.Overall, there is a general increase of mixed-types over time, which is expected given the rise of global agendas understanding interconnections of social, environmental, and economic domains such as the Millennium Development Goals and the Sustainable Development Goals.The most frequent mixed-type problem construction of the Amazon is economic conservation. This mixed-type construction, composed of Amazonian statements that construct the Amazon as a problem of both economic integration and environmental conservation, generally increased over time. This increase suggests that later presidents, along with being more diverse in how they construct the Amazon as an issue, increasingly find ways to reconcile the dictum between the economy and the environment that had prevailed in the previous years.Mixed-types are particular interesting considering both Lula and Bolsonaro, who construct the Amazon as a multifaceted issue more frequently than other presidents. Lula usually mixed economic integration with environmental conservation and social development when constructing the Amazon as an issue, while Bolsonaro constructs the Amazon as an issue of national sovereignty and economic integration much more than any other president, mimicking the military dictatorship discourses and policies toward the region [@hecht1990]. -->

@@

By adapting transnational networks to discourse, we argue that presidents' expectations about transnational audiences in different settings explain variations in problem-constructions. To investigate the relationship between where presidents speak and how they construct the Amazon as an issue, we run four logistic regressions, one for each pure-type construction as a dependent variable (see table 2 below). Multiple, often contradictory, problem-constructions are forwarded across settings as presidents participate and shape transnational politics, confirming what our conceptual frameworks foresee. We discuss three main findings:

First, when speaking to international audiences (be it in governmental summits, bilateral visits, or fora) presidents are between 54.6% and 56.1%[2] more likely to construct the Amazon as an issue of environmental conservation, in relations to when they speaking inside Amazonian states within Brazil (reference category). When speaking within Amazonian countries and Amazonian states, our models suggest that the likelihood of deploying environmental conservation decreases drastically. We interpret this as evidence that presidents construct the Amazon as an environmental problem outside Brazil, because of their views of transnational audiences as possible funders of domestic policy and projects. The Brazilian government successfully funded domestic public policy with transnational support on several occasions: the 1992 Pilot Programme of the G7 for the Protection of Rainforest; the 2001 Amazon Region Protected Areas Program by the World Bank and others; the 2008 Amazon Fund which provided over USD 1 billion; multiple Global Environment Facility grants; as well as multiple bilateral and philanthropic grants [@silva-muller2022]. They realize material and symbolic gains (leverage politics) by making use of information that argues in favor of the Amazon as an environmental problem (information politics).

Second, economic integration and social development are more likely inside the Amazon than in an international setting. Concurrent with the successful environmental policies from the early 2000s, financed by transnational monies, presidents also pursued a national agenda of economic integration or social development. For instance, rural credit offered to local agricultural producers in Amazonian states, went up from 500 million reais a year in 1999 to over 4 billion a year by 2012 [@capobianco2021]. The special fiscal regime in Manaus, designed to increase industry presence in the region, was maintained. More roads and dams (e.g. BR-163 and Belo Monte) were constructed during the period. Presidents likely highlight these policies as they see Amazonian populations as voters to whom they are accountable. We argue that presidents expect voters' to have socio-economic interests, rather than environmental ones, and therefore construct the Amazon accordingly.

@@

When it comes to sovereignty, model four also reveals that sovereignty seems to be a problem construction used more within the country than outside. Interestingly, while we found that the frequency of the Amazon as a general topic is driven by environmental-related events (section 4.1), presidents do not necessarily respond to them with environmental problem-constructions. Rather, as such events bring strong international media coverage, Brazilian presidents match them with discursive attempts to highlight Brazil's sovereignty over the Amazonian territory. The 2019 August Fires are exemplary of these dynamics: in a series of responses directed to President Macron, Bolsonaro consistently highlighted Brazil's sovereignty in various speeches inside Brazil.

@@

Finally, while our models take settings and policy outcomes (deforestation rates) as cummulative, we use figure 5 to reveal a complex relationship between policy outcomes, problem constructions and settings. We focus on problem-constructions over time inside Amazonian States and in International settings for simplicity. Figure 5 demonstrates that when deforestation rates improved (2004-2012),presidents preferred deploying environmental conservation both inside the Amazon as well as outside Brazil. As they worsen(2012-2021), the environmental conservation problem-construction stays steady inside Brazil, but decreases strongly outside. In line with our framework, this suggests presidents boast good outcomes whenever possible by highlighting the aligned problem construction. When outcomes are bad, we see other problem constructions increasing instead, likely deployed as means of sidestepping issues.

Overall, we find that presidents promote economic integration and social development within Brazil that diverges from the environmental ones promoted outside of Brazil. They talk to the economic or social needs of the voters domestically, while boasting about environmentalism internationally. Sovereignty concerns, while closely tied to international politics, are more frequent inside the country. Two-level game theories cannot account for these variations. Rather, these results point to presidents as participants of transnational networks: they speak based on what they can expect of an audience given the context.

@@

This paper investigates how the Brazilian Amazon has been constructed as a problem in Brazilian presidential speeches since 1985. We have four main findings: first, the frequency of the Amazon as a general topic in speeches is driven by environmental related events that draw international attention. Socioeconomic politics in the region, contrarily, does not seem to draw attention or drive presidents to speak more. Second, we demonstrate how the same government adopts multiple problem-construction concurrently. The expectations of presidents about the audience of the speech, more than anything else, is the driver of variation of problem-construction. Third, and relatedly, we identify a long-term reduction trend of economic integration to a more balanced deployment of all problem-constructions. Finally, we demonstrate how domestic outcomes such as deforestation drive specific problem-constructions cunningly: they can be used to side-step problems to a specific audience.

>>>>

## 3.1 Amazonian historiography

To capture the cultural dimension and historical contexts behind problem constructions, we rely on the Amazonian historiography literature. We refer to Amazonian historiography as the body of research conducted by social and environmental scientists that tell the story of diverse governmental programs that studies the Amazon region, forest, and its people. Within this literature, we identify three historical-political constructions: national sovereignty, economic integration, and environmental conservation. Based on an inductive analysis of our empirical material, we argue that a fourth separate construction also exists: social development. The social development construction focuses on the constitutional rights and dignity of the inhabitants of the Amazonian region.

@@

We code Amazonian statements in three steps. First, we develop a codebook for coding Amazonian problem-constructions based on the Amazonian historiography (see Codebook in appendix). In their conceptualization, each problem-construction is mutually exclusive; that is, they cover different forms of constructing the Amazon as a problem. However, an Amazonian statement can be coded as referring to multiple problem constructions (see Figure 1 below). Amazonian statements, thus, can be pure-type statements (i.e. constructs the Amazon as a single problem) or mixed-types (i.e. constructs the Amazon as multiple issues).

@@

To analyze our data, we first visualize when presidents speak more about the Amazon and how they construct the Amazon as a problem in time. We then model the share of a specific pure problem-construction, as the dependent variable, relative to the settings in which a particular speech happens, as our independent variable. This means we predict the probability of a president constructing the Amazon as an issue of (1) environmental conservation, (2) economic integration, (3) social development, (4) national sovereignty, separately, and in relation to the remaining problem-constructions (see table 1 below). We focus on single problem constructions to provide unambiguous interpretations and more robust findings. We divide settings in terms of non-Amazonian states, Amazonian states, Amazonian countries, and non-Amazonian countries. This level of aggregation refers to geo-political settings that make sense in light of the Amazon, as our policy object. This aggregation combines different expectations presidents can have about transnational audiences in each of these settings. The model controls for yearly deforestation rates, inflation, election years, and peak years for Amazon in presidential speeches.

@@

Second, our dataset covers only what is considered an official speech. Presidents, though, give interviews, appear in debates, talk at campaign rallies, and, more recently, post on social media. Problem-construction within presidential discourse, thus, happens in different public, private, and online settings for which we do not account or differentiate in this paper. Furthermore, we aggregate levels to socio-political-cultural settings that make sense in light of the policy object in question and the data at hand. Therefore, we are aware that a category as "non-Amazonain countries", for instance, encompasses speeches delivered by presidents in a wide variety of countries, institutions, and venues to diverse audiences which are not differentiated.

Third, as presidents participate and shape transnational politics, we assume that presidential discourse, policies, and outcomes are co-constitutive. Discourses capture general problem constructions that can be used to introduce, discuss, and/or justify policy and outcomes. Throughout the analysis we carefully discuss how presidential discourses could relate to certain outcomes by pointing at the patterns but without making causality claims.

@@

We subsequently observe a steady increase reaching 24% in 2019, the first year of Bolsonaro's presidency. At the time, international and national media brought attention to record burning of the Amazon, which led up to a red sky afternoon in São Paulo in August 2019. Alongside this, Bolsonaro's dismantling of environmental governance and the threats to leave the Paris Agreement had strong media attention, and, as a response, the topic became prominent in his speeches. Bolsonaro went on to retrieve Brazil's hosting status for COP25. Finally, and unfortunately, in 2022, environmental activist, Bruno Mendes, and British journalist, Dom Phillips, were murdered and buried in the Amazon (McCoy 2022).
<!-- Discourses are also a response to political pressures; are not just problem-constructions but also a sort of justification... (Grazi's comment) -->

@@

Although we observe variation in the frequency of the Amazon as a general topic, our conceptual framework foresees variation in specific problem-construction depending on a myriad of factors. In this subsection, we investigate how pure-types and mixed-types problem constructions evolved over time. The same president simultaneously constructs the Amazon as various problems, which unravels a less cohesive story than that told by the policy-cycle literature. Nevertheless, the overall balance between problem-constructions reveals longitudinal trends that speak to the importance of the environment, economy, social policy, and sovereignty. Figure 3, below, illustrates the share of each pure-type problem-constructions, along with the share of all mixed-types together, in time.

@@

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso's presidency (1995-2002), when economic integration was increasing while environmental conservation decreased. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. While this correlates with strong environmental policies being adopted in 2004, economic integration still ranked first. It was only in 2010, with President Dilma, that environmental conservation and social development problem-constructions surpassed economic integration. Interestingly, there is a steady increase of sovereignty problem-constructions starting in 2008 already. This reversal was not long-standing, though. With the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration increases to become the most favored construction.

The same governments deploy contradictory constructions, but their relative frequencies reveal trends over time. The 1988 Brazilian Constitution foresaw the strengthening of environmentalism through the participation of indigenous peoples and civil society in policy-making [@keck1998], which was slowly confirmed, at least discursively. The fall of economic integration problem-construction and the rise of both social development and environmental conservation constructions in the 2000s suggest an unprecedented balance between granting local livelihoods their rights and economic development: never, since 1985, Amazonian constructions that focused on the environment and the constitutional rights of its people were more present than the economic integration in presidential discourses. While social development and environmental conservation did go down, we are now at a point in time where they appear almost as frequently as economic integration.

The steady increase of sovereignty as salient problem-constructions starts in 2010, much before Bolsonaro was elected. The revision of the Forest Code, adopted in 2012, inaugurated a new phase of strong representation of agri-businesses in government and congress and, consequently, strong civil society criticism of government. In a period when sovereignty-related issues broadened from nation-states to a wider set of actors, stronger criticism of civil society in politics was often met by presidents with accusations of internationalization attempts. Thus, the political forces in Brazilian politics that drove policy and discursive changes in problem-construction were long in the making. Bolsonaro's problem-constructions, characterized by a unique combination of economic integration and sovereignty, is the strongest form of this shift.

Finally, our operationalization foresees the possibility of presidents might mixing multiple problem-constructions within a same statement about the Amazon. Mixed-type problem-constructions in discourse offer more intricate understandings of the Amazon as a problem. Constructing the Amazon as multiple issues increases from 18%, in 1985, to 29%, in 2022, of all constructions. This general increase is expected given the rise of global agendas understanding interconnections of social, environmental, and economic domains such as the Millennium Development Goals and the Sustainable Development Goals. This increase suggests that later presidents, along with being more diverse in how they construct the Amazon as an issue, increasingly find ways to reconcile the dictum between the economy and the environment that had prevailed in the previous years.

@@

By adapting transnational networks to discourse, we argue that presidents' expectations about transnational audiences in different settings explain variations in problem-constructions. To investigate the relationship between where presidents speak and how they construct the Amazon as an issue, we run five logistic regressions, one for each pure-type construction as a dependent variable and one for all mixed-type constructions (see table 2 below). Multiple, often contradictory, problem-constructions are forwarded across settings as presidents participate and shape transnational politics, confirming what our conceptual frameworks foresee.

@@

On the one hand, when speaking at non-Amazonian countries (be it in governmental summits, bilateral visits, or fora), presidents are more likely to construct the Amazon as an issue of environmental conservation, in relation to when they speaking within Amazonian states in Brazil (reference category). To a lesser extent, this also holds for when presidents speak in non-Amazonian states in Brazil. Alternatively, when speaking within Amazonian countries, the likelihood of deploying environmental conservation decreases drastically. We interpret this as evidence that presidents often construct the Amazon as an environmental problem outside of Amazonian states within Brazil, and of Amazonian neighboring countries, because of their views of transnational audiences as possible funders of domestic policy and projects. for instance, the Brazilian government successfully funded domestic public policy with transnational support on several occasions: the 1992 Pilot Programme of the G7 for the Protection of Rainforest; the 2001 Amazon Region Protected Areas Program by the World Bank and others; the 2008 Amazon Fund which provided over USD 1 billion; multiple Global Environment Facility grants; as well as multiple bilateral and philanthropic grants [@silva-muller2022]. That is, they realize material and symbolic gains internationally, outside of Amazonian countries, by constructing the Amazon as an environmental problem.

On the other hand, economic integration problem constructions are more likely to take place within Amazonian countries than within Amazonian states. Whereas social development constructions are less likely to take place within Amazonian countries and non-Amazonian countries. Concurrent with the successful environmental policies from the early 2000s, financed by transnational monies, presidents also pursued a national agenda of economic integration or social development. For instance, rural credit offered to local agricultural producers in Amazonian states, went up from 500 million reais a year in 1999 to over 4 billion a year by 2012 [@capobianco2021]. The special fiscal regime in Manaus, designed to increase industry presence in the region, was maintained. More roads and dams (e.g. BR-163 and Belo Monte) were constructed during the period. Presidents likely highlight these policies as they see Amazonian populations as voters to whom they are accountable. We argue that presidents expect voters' to have more immediate socio-economic interests, rather than environmental ones, and therefore construct the Amazon accordingly.

Sovereignty problem constructions, interestingly, are more likely to take place when presidents speak at non-Amazonian states in Brazil or when relevant environmental-related events occur. While we found that the frequency of the Amazon as a general topic is driven by environmental-related events (section 4.1), presidents do not necessarily respond to them with specific problem-constructions. Rather, as such events bring strong international media coverage, Brazilian presidents match them with discursive attempts to highlight Brazil's sovereignty over the Amazonian territory.

Furthermore, constructions of the Amazon as a problem of economic integration increase as deforestation rates also increase; while constructions of the Amazon as a problem of environmental conservation increase as deforestation rates decrease. Although our models take settings and policy outcomes without differentiating for their different correlations as they vary in time, there is a complex relationship between problem constructions, settings, and time. For instance, figure 5, below, demonstrates that when deforestation rates decreased form 2004 to 2012, presidents preferred constructing the Amazon as an issue of environmental conservation both within Amazonian states and in non-Amazonian countries. As deforestation rates increased from 2012 to 2021, environmental conservation problem-construction remained steady within Amazonian states but decreased sharply in non-Amazonian countries. Alternatively, the positive relationship between economic integration problem constructions and deforestation rates from the model is driven by the same relationship within Amazonian states. When speaking at non-Amazonian countries, presidents are generally less likely to construct the Amazon as an issue of economic integration in time. In line with our framework, this suggests presidents boast positive policy outcomes whenever possible and wherever profitable by leveraging the aligned problem construction. Alternatively, presidents sidestep constructions whenever policy outcomes might be negative and wherever unlikely that the aligned problem-construction will be transformed in leverage.

@@

Overall, we find that presidents promote economic integration and social development within Brazil and neighboring Amazonian countries that diverges from the environmental one promoted non-Amazonian countries. That is, presidents might talk to the economic or social needs of the voters domestically, while boasting about environmentalism internationally. Presidents, as participants of transnational networks, speak based on what they can expect of an audience given the context.

@@

This paper investigates how the Brazilian Amazon has been constructed as a problem in Brazilian presidential speeches since 1985. We have four main findings: first, the frequency of the Amazon as a general topic in speeches is driven by environmental related events that draw international attention. Second, we demonstrate how the same government adopts multiple problem-construction concurrently. The expectations of presidents about the audience of the speech, more than anything else, is the driver of variation of problem-construction. Third, and relatedly, we identify a long-term reduction trend of economic integration to a more balanced deployment of all problem-constructions. Finally, we demonstrate how presidents promote economic integration and social development within Brazil and neighboring Amazonian countries that diverges from the environmental one promoted non-Amazonian countries.
